References for PloS paper

### 2014-01-14 image_analysis_paper/plos_draft.tex
Added more edits to paper

The issue of rotation invariance arises in many applications, including medical imaging \cite{hajnal2010medical}, face recognition \cite{rowley1998rotation}, digit recognition \cite{simard1992efficient}, and texture classification \cite{greenspan1994rotation}.
%
In these settings, objects from different frames or timepoints must be aligned or ``registered'' prior to further analysis.
%
%For example, often medical images (such as MRI scans) are collected over multiple days, and doctors would like to study development of a condition (such as the growth of a tumor).
%
%However, on each day, the patient and/or sensors may be in slightly different locations; therefore, the images must be aligned before the doctor can look at how the tumor is growing and evolving.
%
Several different techniques have been developed to solve the image registration problem. 
%
``Template-based'' techniques, where one selects a prototypical image or ``template'' and then aligns the remaining images to this template, are relatively common \cite{ahuja2007template}.
%
Such techniques can suffer if a good template is not know {\em a priori}; if the images are noisy, selecting one image from the data set at random can yield a noisy template and result in many mis-alignments. 
%
Other techniques focus on constructing ``features'' that are rotation-invariant \cite{flusser2000independence, lowe1999object, sadler1992shift, ojala2002multiresolution}, so that one can use these features (rather than the images themselves) in further analyses and computations.
%
However, developing rotation-invariant features that still retain the important information in the images is difficult and often problem-dependent, and constructing features that are stable to noise and deformations is still an active area of research \cite{mallat2012group, sifre2013rotation}.
%
One can also construct algorithms that consider all possible image rotations in the analysis \cite{hilai1994recognition, zhao2013fourier}.
%
%TODO: why do these methods fail?

We are interested in analyzing images of developing {\em Drosophila} embryos during nuclear cycle 14 (NC14).
%TODO: check that NC14 is correct
%
More specifically, we would like to order the images {\em in time} so that we can reconstruct the developmental dynamics from snapshots. 
%
% TODO: add motivation about why we are interested in this
%
Currently, data is collected using a high-throughput microfluidic device \cite{chung2010microfluidic}, so we can image hundreds of embryos in one experiment.
%
Each embryo is stained for the membrane proteins, dpERK, and DI, and we collect fluorescent images of the cross-section of each embryo using a confocal microscope.
%
We then use image analysis software to extract the dpERK concentration profile around the circumfrence of the embryo from the fluorescent image.
%
We use the expression of DI to find the dorsal midline of the embryo, and ``unzip'' the circular profile at the dorsal midline to obtain a one-dimensional, linear dpERK concentration profile.
%
We also measure the membrane thickness of the embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{lim2013kinetics, lecuit2002slam}. 
%
We then order the dpERK concentration profiles for each embryo by the thickness of the membrane to obtain a ``trajectory'' of the dpERK dynamics.

@@

Here,  we will discuss alternative methodologies to reconstruct the dpERK dynamics from fluorescent images. 
%
We will use various dimensionality techniques to {\em automatically} order the dpERK concentration profiles in time.
%
We assume that the data is approximately one-dimensional, and that this one dimension parameterizes time, so that reducing the data to one-dimension and ordering it along this dimension will (approximately) order it in time. 
%
We will not only focus on the linear concentration profiles extracted from the images, but also work directly with the two-dimensional dpERK fluorescent images. 
%
Finally, we will look at ordering the two-dimensional membrane images in time; these are much more feature-rich than the dpERK images, and prove to be more challenging.

@@

We first used principal components analysis (PCA) to automatically order the one-dimensional concentration profiles in time. 
%
We computed the first principal component, $\psi_1$, from the data set, and ordered the data by the projection coefficients onto $\psi_1$.
%
The results are shown in Figure \ref{fig:PCA_ordering}.
%
There is a strong correlation between the first projection coefficient, $\langle x_i, \psi_1 \rangle$, and the membrane thickness;
the Spearman correlation between the first projection coefficient and the membrane thickness is 0.9244.

@@

We would like to note that, because the embryos can be oriented with either the anterior or posterior facing up in the microfluidic device (see Figure \ref{subfig:imaging_device}), the directionality of the concentration profiles (clockwise versus counterclockwise around the ring, or right-to-left versus left-to-right for the linear profiles) is not meaningful. 
%
Therefore, we could do PCA with an expanded data set where we also include the mirror image of each concentration profile.
%
Symmetries in PCA have been studied extensively in previous work \cite{holmes1998turbulence}; it was shown that when mirror images are also included in the data set, the first PCA mode captures the symmetric portion of the variability, and the second mode captures the anti-symmetric portion.
%
We found that including the mirror images of each data point did not change our results significantly, since the first principal component (not shown) is nearly symmetric with respect to left-right reflections.

@@

We have demonstrated that we can order the concentration profiles using PCA or DMAPS.
%
However, ordering the profiles first requires that we factor out the relevant symmetries in the problem.
%
The developmental dynamics are invariant to the orientation of the embryo in the microfluidic device; therefore, to obtain meaningful results from our analysis, we must first align the embryo snapshots.
%
Rotations of the embryo cross-sections correspond to translations (with periodic boundary conditions) of the linear profiles we have been studying.
%
In the previous analyses, we have assumed that the profiles have been prealigned using the location of the dorsal protein peak.
%
However, without some {\em a priori} knowledge of the system (e.g. without knowing that dorsal is expressed at one position along the perimeter of the embryo), we would need to align the concentration profiles before doing our analysis.
%TODO: add when this might be relevant, e.g. when there is no Dorsal

The most common and perhaps intuitive techniques for alignment are template-based methods \cite{ahuja2007template}, where one chooses are representative data point or shape as a ``template'', and then shifts or rotates each data point to be optimally aligned with this template.
%
Although these methods are simple, they require knowledge of an appropriate template function and can suffer when the data set is noisy \cite{sonday2013noisy}.
%
Angular synchronization \cite{singer2011angular} is a relatively new technique that is often more robust than template-based methods in the presence of noise.
%
We used angular synchronization to factor out the shifts in the concentration profiles measured from the fluorescent images.
%
After aligning the data points using angular synchronization, we used DMAPS to order the aligned concentration profiles.
%
The results are shown in Figure \ref{fig:angsynch_ordering}.
%
One can see that the results are visually consistent with the results shown in Figure \ref{fig:DMAPS_ordering}; the Spearman correlation between $\phi_2$ and the membrane length is 0.9164.

@@

Instead of using angular synchronization followed by DMAPS, we can do the alignment and ordering in one computation using vector diffusion maps (VDM).
%
VDM is similar to angular synchronization, but it operates on data sets that contain both underlying symmetries and dynamics.
%
The results of aligning {\em and} ordering the data using VDM are shown in Figure \ref{fig:vdm_ordering}; the results are similar to those from angular syncrhonization (Figure \ref{fig:angsynch_ordering}), and the Spearman correlation between $\langle \phi_3, \phi_1 \rangle$ and the membrane length is 0.9123.

@@

Thus far, we have focused on the one-dimensional linear concentration profiles that are extracted from the fluorescent images.
%
However, the techniques we have outlined can also be applied to the two-dimensional fluorescent images. 
%
We used VDM to align and order the dpERK fluorescent images. 
%
We now must factor out translations and rotations in the images. 
%
We first projected the images onto a sphere;
rotations and translations of a two-dimensional image then correspond to (three-dimensional) rotations of the sphere.
%
We then used VDM to align and order the (projected) images, with the underlying symmetry group $SO(3)$. 
%
The results are shown in Figure \ref{fig:vdm_image_ordering};
the orderings are of comparable quality to those obtained with the one-dimensional profiles, and the dpERK peaks are nicely aligned.
%
The Spearman correlation between $\langle \phi_4, \phi_1 \rangle$ and the membrane length is 0.9051.

@@

An alternative to aligning the data (as in angular synchronization or VDM) is to construct {\em features} that are invariant to the underlying symmetry group, and then compute distances between features of the data for use in a DMAPS calculation.
%
In our setup, we require features of two-dimensional images that are invariant to translations and rotations of the images.
%
We chose to use the scattering transform \cite{mallat2012group} to construct our invariant features; the scattering transform utilizes a series of wavelet transforms, which are known to perform well for images \cite{akansu2010emerging}.
%
Not only is this transform invariant to translations and rotations, and is also contractive (so that noise and small deformations are not amplified by the transform) \cite{bruna2011classification}.

@@

We hope that we have demonstrated the utility of dimensionality reduction techniques in dynamical systems with underlying symmetries.
%
The work here is intended to provide an overview of the different techniques available, as well as the benefits and shortcomings of each method. 
%
The techniques here were presented from least complex (PCA) to most complex (scattering transforms).
%
More complex techniques allowed us to deal with more complex data, and allowed us to circumvent much of the preprocessing described in Figure~\ref{fig:background}.

The more complex techniques that operate on two-dimensional images do come at a higher computational cost. 
%
The computational time for analyzing the one-dimensional profiles using any of the techniques described (PCA, DMAPS, angular synchronization, VDM) is less than one second. 
%
In contrast, the analysis of two-dimensional images using either VDM or the scattering transform with DMAPS is $\mathcal{O}(100)$ seconds.
%
The expensive portion of the VDM calculation is computing the pairwise alignments of the images; the expensive portion of the scattering transform with DMAPS is computing the scattering transform coefficients.
%
However, computing the pairwise alignments is $\mathcal{O}(n^2)$, where $n$ is the number of data points, while computing the scattering transform coefficients is $\mathcal{O}(n)$.
%
We therefore expect for VDM to be computationally intractable for large data sets.
%
The added computational cost of VDM is not without benefits, however; VDM produces optimal alignments for each image (in addition to an ordering), while the scattering transform gives us no alignments.

We would like to note that our system was such that the data was inherently one-dimensional, and that this one dimension was monotonic in time.
%
However, this monotonicity was a special feature of our system and is not guaranteed in general.

@@

\subsection*{Image preprocessing}
The dpERK images we collected from the confocal microscope were subsampled to $100 \times 100$ pixels; 
the membrane images were subsampled to $200 \times 200$ pixels (due to the finer structure in the membrane pictures). 
%
Because we are only interested in the membrane thickness, we ran a simple edge detection algorithm on the membrane images.
%
We used a Laplacian of Gaussians filter \cite{canny1986computational}, as implemented in the Matlab Image processing toolbox via the \texttt{edge} function (\url{http://www.mathworks.com/help/images/ref/edge.html}).
%
To compute the pairwise alignments between images for use in angular synchronization and VDM, we searched over 20 rotations (evenly spaced from $0$ to $2 \pi$) and 10 translations spanning $\pm 10 \%$ of the image pixels.
%
We used the Euclidean distance between images to measure the quality of the alignment.
%
These translations and rotations were then converted to three-dimensional rotations on the sphere, with the image spanning a $\pi/4 \times \pi/4$ portion of the sphere, so that the angle of rotation of the image corresponded to the Euler angle $\alpha$, $pixels_x/pixels_{max} * \pi/4$ corresponded to the Euler angle $\beta$, and  $pixels_y/pixels_{max} * \pi/4$ corresponded to the Euler angle $\gamma$.

@@

Spearman's rank correlation coefficient, or Spearman's $\rho$, measures the monotonicity of a function \cite{myers2010research}. 
%
It therefore serves as an appropriate metric for the ``goodness'' of our orderings.
%
Given data points $X_i$ and $Y_i$, we compute the ranks of each data point within the data set, denoted $x_i$ and $y_i$.
%
Spearman's $\rho$ is then defined as the correlation between the ranks
\begin{equation}
\rho = \frac{\sum_i (x_i - \overline{x})(y_i - \overline{y})}{\sqrt{\sum_i (x_i - \overline{x})^2 \sum_i (y_i - \overline{y})^2}}
\end{equation}
where $\overline{x}$ and $\overline{y}$ are the averages of $x_i$ and $y_i$, respectively.

@@

Let $X \in \mathbb{R}^{n \times d}$ denote our data set consisting of $n$ points in $d$ dimensions, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
%
We first mean-center the data set to produce the matrix $\hat{X}$, where
\begin{equation}
\hat{X}_{ij} = X_{ij} - \frac{1}{n} \sum_{k=1}^n X_{kj}.
\end{equation}
%
The empirical covariance of the data set, $C \in \mathbb{R}^{d \times d}$ is then given by
\begin{equation}
C = X^T X.
\end{equation}
%
We compute the eigenvectors $\psi_1, \dots, \psi_d$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ of $C$, and order them such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d$.
%
The principal components of the data set $X$ are given by $\psi_1, \dots, \psi_d$, and the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ measure the variability captured by the corresponding eigenvector. 
%
If there is a {\em spectral gap} such that the first few eigenvectors capture a significant portion of the variance in our data set, projection onto these first few principal components will retain most features of the data set while reducing the dimensionality.
%
In our setup, we assume that the data is inherently one-dimensional, and that projection onto $\psi_1$ will uncover the correct time ordering of the data.

\subsection*{Diffusion maps}
Unlike PCA, diffusion maps (DMAPS) is a nonlinear dimensionality reduction technique. 
%
DMAPS aims to uncover the parameterization of data sampled from a nonlinear manifold \cite{coifman2005geometric}.
%
The {\em essential} requirement for DMAPS is an appropriate distance metric $d(x_i, x_j)$ for comparing data points. 
%
This can as simple as the standard Euclidean distance, or a more complex metric (such as a distance between features of the data points) for other data sets.

@@

When the data set is invariant under an underlying symmetry group (such as translations and/or rotations), we must first factor out the underlying symmetries before doing further analysis.
%
Angular synchronization calculates the optimal alignments for a data set, assuming we have estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.

>>>>

The issue of rotation invariance arises in many applications, including face recognition \cite{rowley1998rotation}, medical imaging \cite{hajnal2010medical}, digit recognition \cite{simard1992efficient}, and texture classification \cite{greenspan1994rotation}.
%
In these settings, images from different timepoints, users, or contexts must be aligned or ``registered'' prior to further analysis.
%
%For example, often medical images (such as MRI scans) are collected over multiple days, and doctors would like to study development of a condition (such as the growth of a tumor).
%
%However, on each day, the patient and/or sensors may be in slightly different locations; therefore, the images must be aligned before the doctor can look at how the tumor is growing and evolving.
%
Several different techniques have been developed to solve the image registration problem. 
%
Template-based techniques, where one selects a prototypical image or ``template'' and then aligns the remaining images to this template, are relatively common \cite{ahuja2007template}.
%
Such techniques can suffer if a good template is not know {\em a priori}; if the images are noisy, selecting one image from the data set at random can yield a noisy template and result in many mis-alignments. 
%
Other techniques focus on constructing ``features'' that are rotation-invariant \cite{flusser2000independence, lowe1999object, sadler1992shift, ojala2002multiresolution}, so that one can use these features (rather than the images themselves) in further analyses and computations.
%
However, developing rotation-invariant features that still retain the important information in the images is difficult and often problem-dependent, and constructing features that are stable to noise and deformations is still an active area of research \cite{mallat2012group, sifre2013rotation}.
%
One can also construct algorithms that consider all possible image rotations in the analysis \cite{hilai1994recognition, zhao2013fourier}.
%
While these techniques are often robust to additive Gaussian noise, they are often sensitive to deformations and other types of noise typically encountered in images.

We are interested in analyzing images of developing {\em Drosophila} embryos during nuclear cycle 14 (NC14).
%TODO: check that NC14 is correct
%
More specifically, we are interested in the activation dynamics of dpERK during NC14.
%
We collect fluorescent images of {\em Drosophila} embryos that have been stained for dpERK. 
%
Because we must fix the embryos before staining for dpERK, we cannot obtain a continuous time trajectory of dpERK in one embryo; instead, we obtain a set of snapshots, one per embryo, of embryos stained for dpERK. 
% 
We would like to order the resulting images {\em in time} so that we can reconstruct the developmental dynamics from snapshots. 
%
% TODO: add motivation about why we are interested in this
%
Currently, data is collected using a high-throughput microfluidic device \cite{chung2010microfluidic}, so we can image hundreds of embryos in one experiment.
%
Each embryo is stained for the membrane proteins, dpERK, and DI, and we collect fluorescent images of the cross-section of each embryo using a confocal microscope.
%
We then use image analysis software to extract the dpERK concentration profile around the circumfrence of the embryo from the fluorescent image.
%
We use the expression of DI to find the dorsal midline of the embryo, and ``unzip'' the circular profile at the dorsal midline to obtain a one-dimensional, linear dpERK concentration profile.
%
We also measure the membrane thickness of the embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{lim2013kinetics, lecuit2002slam}. 
%
We then order the dpERK concentration profiles for each embryo by the thickness of the membrane to obtain a ``trajectory'' of the dpERK dynamics.

@@

Here,  we will discuss alternative methodologies to reconstruct the dpERK dynamics from fluorescent images. 
%
%TODO: add why we cannot always do the procedure outlined above
We will use various dimensionality techniques to {\em automatically} order the dpERK concentration profiles in time.
%
We assume that the data is approximately one-dimensional, and that this one dimension parameterizes time, so that reducing the data to one dimension and ordering it along this dimension will (approximately) order it in time. 
%
We will not only focus on the linear concentration profiles extracted from the images, but also work directly with the two-dimensional dpERK fluorescent images. 
%
Finally, we will look at ordering the two-dimensional membrane images in time; these are much more feature-rich than the dpERK images, and prove to be more challenging.

@@

We first used principal components analysis (PCA) to order the one-dimensional concentration profiles in time. 
%
We computed the first principal component, $\psi_1$, from the data set of concentration profiles, and ordered the data by the projection coefficients onto $\psi_1$.
%
The results are shown in Figure \ref{fig:PCA_ordering}.
%
There is a strong correlation between the first projection coefficient, $\langle x_i, \psi_1 \rangle$, and the membrane thickness;
the Spearman correlation between the first projection coefficient and the membrane thickness is 0.9244.

@@

We would like to note that, because the embryos can be oriented with either the anterior or posterior facing up in the microfluidic device (see Figure \ref{subfig:imaging_device}), the directionality of the concentration profiles (clockwise versus counterclockwise around the ring, or right-to-left versus left-to-right for the linear profiles) is not meaningful. 
%
Therefore, we could do PCA with an expanded data set where we also include the mirror image of each concentration profile.
%
Symmetries in PCA have been studied extensively in previous work \cite{holmes1998turbulence}; it was shown that when mirror images are also included in the data set, the first PCA mode captures the symmetric portion of the variability, and the second mode captures the anti-symmetric portion.
%
We found that including the mirror images of each data point did not change our results significantly, since the first principal component computed using the original data set (not shown) is nearly symmetric with respect to left-right reflections.

@@

We have demonstrated that we can order the concentration profiles using PCA or DMAPS.
%
However, ordering the profiles first requires that we factor out the relevant symmetries in the problem.
%
The developmental dynamics are invariant to the orientation of the embryo in the microfluidic device; therefore, to obtain meaningful results from our analysis, we must first align the concentration profiles.
%
Rotations of the embryo cross-sections correspond to translations (with periodic boundary conditions) of the linear profiles we have been studying.
%
In the previous analyses, we have assumed that the profiles have been prealigned using the location of the dorsal protein peak.
%
However, without some {\em a priori} knowledge of the system (e.g. without knowing that dorsal is expressed at one position along the perimeter of the embryo), we would need to align the concentration profiles before doing our analysis.
%TODO: add when this might be relevant, e.g. when there is no Dorsal

The most common and perhaps intuitive techniques for alignment are template-based methods \cite{ahuja2007template}, where one chooses are representative data point or shape as a ``template'', and then shifts or rotates each data point to be optimally aligned with this template.
%
Although these methods are simple, they require knowledge of an appropriate template function and can suffer when the data set is noisy \cite{sonday2013noisy}.
%
Angular synchronization \cite{singer2011angular} is a relatively new technique that is often more robust than template-based methods in the presence of noise.
%
We used angular synchronization to factor out the shifts and align the concentration profiles measured from the fluorescent images.
%
Because the concentration profiles lie on a circle (we unwrap them onto a line for visualization purposes), factoring out shifts of the linear profiles (with periodic boundary conditions) corresponds to factoring out rotations of a circle with the underlying symmetry group $SO(2)$. 
%
After alignment, we used DMAPS to order the aligned concentration profiles.
%
The results are shown in Figure \ref{fig:angsynch_ordering}.
%
One can see that the results are visually consistent with the results shown in Figure \ref{fig:DMAPS_ordering}; the Spearman correlation between $\phi_2$ and the membrane length is 0.9164.

@@

Instead of using angular synchronization followed by DMAPS, one can do the alignment and ordering in one computation using vector diffusion maps (VDM).
%
VDM is similar to angular synchronization, but it operates on data sets that contain both underlying symmetries and dynamics.
%
The results of aligning {\em and} ordering the data using VDM are shown in Figure \ref{fig:vdm_ordering}; the results are similar to those from angular syncrhonization (Figure \ref{fig:angsynch_ordering}), and the Spearman correlation between $\langle \phi_3, \phi_1 \rangle$ and the membrane length is 0.9123.

@@

Thus far, we have focused on the one-dimensional linear concentration profiles that are extracted from the fluorescent images.
%
However, the techniques we have outlined can also be applied to the two-dimensional fluorescent images. 
%
We used VDM to align and order the dpERK fluorescent images. 
%
We now must factor out translations and rotations in the images. 
%
We first projected the images onto a sphere;
rotations and translations of a two-dimensional image then correspond to (three-dimensional) rotations of the sphere.
%
We used VDM to align and order the (projected) images, with the underlying symmetry group $SO(3)$. 
%
The results are shown in Figure \ref{fig:vdm_image_ordering};
the orderings are of comparable quality to those obtained with the one-dimensional profiles, and the dpERK peaks are nicely aligned.
%
The Spearman correlation between $\langle \phi_4, \phi_1 \rangle$ and the membrane length is 0.9051.

@@

An alternative to aligning the data (as in angular synchronization or VDM) is to construct {\em features} that are invariant to the underlying symmetry group, and then compute distances between features of the data for use in a DMAPS calculation.
%
In our setup, we require features of two-dimensional images that are invariant to translations and rotations of the images.
%
We chose to use the scattering transform \cite{mallat2012group} to construct our invariant features; the scattering transform utilizes a series of wavelet transforms, which are known to perform well for images \cite{akansu2010emerging}.
%
Not only is this transform invariant to translations and rotations, but is also a contractive operator \cite{bruna2011classification}, so that noise and small deformations are not amplified by the transform.

@@

We hope we have demonstrated the utility of dimensionality reduction techniques in systems with underlying symmetries and dynamics.
%
The work here is intended to provide an overview of the different techniques available, as well as the benefits and shortcomings of each method. 
%
The techniques were presented from least complex (PCA) to most complex (scattering transforms).
%
More complex techniques allowed us to deal with more complex data, and allowed us to circumvent much of the preprocessing described in Figure~\ref{fig:background}.

The more complex techniques that operate on two-dimensional images do come at a higher computational cost. 
%
The computational time for analyzing the one-dimensional profiles using any of the techniques described (PCA, DMAPS, angular synchronization, VDM) is less than one second. 
%
In contrast, the analysis of two-dimensional images using either VDM or the scattering transform with DMAPS is $\mathcal{O}(100)$ seconds.
%
The expensive portion of the VDM calculation is computing the pairwise alignments of the images, and the expensive portion of the scattering transform with DMAPS is computing the scattering transform coefficients.
%
However, computing the pairwise alignments is $\mathcal{O}(n^2)$, where $n$ is the number of data points, while computing the scattering transform coefficients is $\mathcal{O}(n)$.
%
We therefore expect for VDM to be computationally intractable for large data sets.
%
The added computational cost of VDM is not without benefits; VDM produces optimal alignments for each image (in addition to an ordering), while the scattering transform gives us no alignments.

We would like to note that our system was such that the data was inherently one-dimensional, and that this one dimension was monotonic in time.
%
However, this monotonicity was a special feature of our system and is not guaranteed in general.
%TODO: add more conclusions

@@

\subsection*{Image preprocessing}
The dpERK images we collected from the confocal microscope were subsampled to $100 \times 100$ pixels; 
the membrane images were subsampled to $200 \times 200$ pixels (due to the finer structure in the membrane pictures). 
%
Because we were only interested in the membrane thickness, we ran a simple edge detection algorithm on the membrane images.
%
We used a Laplacian of Gaussians filter \cite{canny1986computational}, as implemented in the Matlab Image processing toolbox via the \texttt{edge} function (\url{http://www.mathworks.com/help/images/ref/edge.html}).
%
To compute the pairwise alignments between images for use in angular synchronization and VDM, we searched over 20 rotations (evenly spaced from $0$ to $2 \pi$) and 10 translations spanning $\pm 10 \%$ of the image pixels.
%
We used the Euclidean distance between images to measure the quality of the alignment.
%
These translations and rotations were then converted to three-dimensional rotations on the sphere, with the image spanning a $\pi/4 \times \pi/4$ portion of the sphere.%, so that the angle of rotation of the image corresponded to the Euler angle $\alpha$, $pixels_x/pixels_{max} * \pi/4$ corresponded to the Euler angle $\beta$, and  $pixels_y/pixels_{max} * \pi/4$ corresponded to the Euler angle $\gamma$.

@@

Spearman's rank correlation coefficient, or Spearman's $\rho$, measures the monotonicity of a function \cite{myers2010research}. 
%
It therefore serves as an appropriate metric for the quality of our orderings.
%
Given two data sets $X$ and $Y$, with data points $X_i$ and $Y_i$, we compute the ranks of each data point within the data set, denoted $x_i$ and $y_i$.
%
Spearman's rank correlation between $X$ and $Y$ is then defined as the correlation between the ranks
\begin{equation}
\rho = \frac{\sum_i (x_i - \overline{x})(y_i - \overline{y})}{\sqrt{\sum_i (x_i - \overline{x})^2 \sum_i (y_i - \overline{y})^2}}
\end{equation}
where $\overline{x}$ and $\overline{y}$ are the averages of $x_i$ and $y_i$, respectively.

@@

Let $X \in \mathbb{R}^{n \times d}$ denote our data set, consisting of $n$ points in $d$ dimensions, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
%
We first mean-center the data set to produce the matrix $\hat{X}$, where
\begin{equation}
\hat{X}_{ij} = X_{ij} - \frac{1}{n} \sum_{k=1}^n X_{kj}.
\end{equation}
%
The empirical covariance of the data set, $C \in \mathbb{R}^{d \times d}$ is then given by
\begin{equation}
C = X^T X.
\end{equation}
%
We compute the eigenvectors $\psi_1, \dots, \psi_d$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ of $C$, and order them such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d$.
%
The principal components of the data set $X$ are given by $\psi_1, \dots, \psi_d$, and the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ measure the variability captured by the corresponding eigenvector. 
%
If there is a {\em spectral gap} such that the first few eigenvectors capture a significant portion of the variance in our data set, projection onto these first few principal components will retain most features of the data set while reducing the dimensionality.
%
In our setup, we assume that the data is inherently one-dimensional, and that projection onto $\psi_1$ will uncover the correct time ordering of the data.

\subsection*{Diffusion maps}
Unlike PCA, diffusion maps (DMAPS) is a nonlinear dimensionality reduction technique. 
%
DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold \cite{coifman2005geometric}.
%
The {\em essential} requirement for DMAPS is an appropriate distance metric $d(x_i, x_j)$ for comparing data points. 
%
This can as simple as the standard Euclidean distance, or a more complex metric (such as a distance between features of the data points) for other data sets.

@@

When the data set is invariant under an underlying symmetry group (such as rotations), we must first factor out the underlying symmetries before doing further analysis.
%
Angular synchronization calculates the optimal alignments for a data set, assuming we have estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.
